---
title: "WHO_DA 401 Project.Rmd"
author: "Maddy Matura"
date: "February 29, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#Install WHO package created by expersso
install.packages(devtools)
library(devtools)
install_github("expersso/WHO")
```

```{r}
library(WHO)

```




```{r}
#Look at indicators with the word expenditures
expenditures <- codes[grepl("expend", codes$display), ]
```

```{r}
#look at indicators with the word GDP
GDP <- codes[grepl("GDP", codes$display), ]
```

```{r}
#Create GDP dataframe
GDP_df <- get_data("GHED_CHEGDP_SHA2011")
```

```{r}
GHE<- codes[grepl("GHE", codes$display), ] 
```


#Export this data table (GHE per GDP)
```{r}
GHE_percent_GDP <- get_data("GHED_GGHE-DGDP_SHA2011")
```

#Export this data table (Govenrment spending per capita in USD)
```{r}
Spending_per_capita_USD<- get_data("GHED_GGHE-D_pc_US_SHA2011") #Spending per capita in USD for each country
```

```{r}
library(dplyr)
library(tidyr)


```




#Export to excel: GHE percent GDP 2016
```{r}
GHE_percent_GDP16 <- GHE_percent_GDP %>%
  filter(year == "2016")  #filter for 2016
GHE_percent_GDP16 <- subset(GHE_percent_GDP16, select = c(worldbankincomegroup, year,region, country, value))  #select desired columns (income group, year, region, country, value)
GHE_percent_GDP16_noNA <- GHE_percent_GDP16 %>% drop_na()  #get rid of NA
GHE_percent_GDP16_noNA = rename(GHE_percent_GDP16_noNA, GHE_percent = value) #rename columns
```
#Export to excel: Spending per capita 2016
```{r}

Spending_per_capita_USD16 <- Spending_per_capita_USD %>%
  filter(year == "2016")  #filter for 2016 only
Spending_per_capita_USD16<- subset(Spending_per_capita_USD16, select = c(worldbankincomegroup, year,region, country, value)) #subset with desired columns

Spending_per_capita_USD16_noNA <- Spending_per_capita_USD16 %>% drop_na() #drop na

Spending_per_capita_USD16_noNA = rename(Spending_per_capita_USD16_noNA, Spending_per_capita = value)  #rename columns
```

```{r}
Total <- merge(Spending_per_capita_USD16_noNA, GHE_percent_GDP16_noNA, by= "country") #merge dataframes to get total
```

```{r}
Total <- subset(Total, select = c(country, worldbankincomegroup.x, year.x, region.x, Spending_per_capita, GHE_percent)) #pull out desired columns

```


```{r}

Total = rename(Total, worldbankincomegroup = worldbankincomegroup.x)  #rename all columns
Total = rename(Total, year = year.x)
Total= rename(Total, region= region.x)
```



#Mortality
```{r}
Mort <- codes[grepl("[Mm]ortality", codes$display), ]   #search for mortality health indicator
```

```{r}
Mort_df <- get_data("WHOSIS_000004")   #make the mortality indicator into df
```
#Export Adulty Mortality per 1000 people per country in 2016
```{r}
Adult_Mort_2016 <- Mort_df %>% 
  filter(year == "2016", sex== "Both sexes") #filter by 2016 and both sexes


```

```{r}
library(tidyr)
Adult_Mort_2016_noNA <- Adult_Mort_2016 %>% drop_na()  #drop na

Adult_Mort_2016_noNA = rename(Adult_Mort_2016_noNA, Adult_mort = value) #rename columns
```

```{r}
Total <- merge(Total, Adult_Mort_2016_noNA, by= "country") #merge mortality index into total df
```

```{r}
Total <- subset(Total, select = c(country, worldbankincomegroup, year.x, region.x, Spending_per_capita, GHE_percent, Adult_mort))    #select desired columns
```



```{r}  
Total = rename(Total, year = year.x)   #rename
Total= rename(Total, region= region.x)

```

```{r}
Total = rename(Total, Income_Group = worldbankincomegroup) 
```

```{r}
library(ggplot2)
```

```{r}
#Create histogram for total region distribution
hist1 <-ggplot(data=Total, aes(Total$region)) +   
  geom_histogram(stat="count", fill= "red") + labs(x= "Regions", y= "Count", title= "Distribution of Region Counts" ) + theme(plot.title = element_text(hjust = 0.5))
```

```{r}
hist1
```

```{r}
#Create histogram for total income group distribution
hist2 <-ggplot(data=Total, aes(Total$Income_Group)) + 
  geom_histogram(stat="count", fill = "blue") + labs(x= "Income Groups", y = "Count", title = "Distribution of Income Group Counts") + theme(plot.title = element_text(hjust = 0.5))
```

```{r}
hist2
```



```{r}

#Plot spendering per capita by adult mortality index

plot1 <- ggplot(Total,aes( Spending_per_capita, Adult_mort)) + geom_point(aes(col=Income_Group)) + labs(y= "Adult Mortality Rate per 1000 people", x= "Government Healthcare Spending per Capita", title= "Adult Mortality Rate by Government Health Spending per Capita") + theme(plot.title = element_text(hjust = 0.5))
```

```{r}
plot1
```

```{r}
#plot GHE expenditure as a percentage of GDP by adult mortality index
plot2 <- ggplot(Total,aes(GHE_percent, Adult_mort)) + geom_point(col= "red") + geom_smooth(method= "lm") + labs(y="Adult Mortality rate per 1000 people", x="Percent GDP Allocated to Government Health Expenditures", title = "Government Health Expenditures and Adult Mortality Rates") + theme(plot.title = element_text(hjust = 0.5))
```

```{r}
plot2
```

```{r}
#linear model of GHE by adult mortlity index
lm1 <-lm(Adult_mort ~ GHE_percent, data = Total)
```

```{r}
summary(lm1)
```



```{r}
write.csv(Total,"Total.csv")
```



#Health Indices
#All of this is over all ages and all genders
#Physical Burden Index: Necessary Information

#HIV/AIDS
#Deaths per 100,000. The year 2012 is being used because this is the most recent data available

#Adding this in makes the count go down by like 50 :/
#This might be the same situation where some counries are not monitored because the death rate is so low? But I don't know....not sure if I can include this if it takes out so many countries
#OR maybe find a diferent data source
```{r}
HIV<- codes[grepl("HIV", codes$display), ] 
```

```{r}
HIV_DF <- get_data("WHS2_138")


```


```{r}
HIV_DF<-HIV_DF %>%
  filter(year== "2012")
```

```{r}


HIV_DF = rename(HIV_DF, HIV_AIDS = value) #rename columns
```

```{r}
Total2 <- merge(Total, HIV_DF, by= "country") #merge mortality index into total df
```

```{r}
Total2 <- subset(Total2, select = c(country, Income_Group, year.x, region.x, Spending_per_capita, GHE_percent, Adult_mort, HIV_AIDS))    #select desired columns
```



```{r}  
Total2 = rename(Total2, year = year.x)   #rename
Total2= rename(Total2, region= region.x)

```



```{r}
hist3 <-ggplot(data=Total2, aes(Total2$region)) + 
  geom_histogram(stat="count", fill = "blue") + labs(x= "Income Groups", y = "Count", title = "Distribution of Income Group Counts") + theme(plot.title = element_text(hjust = 0.5))
```

```{r}
hist3
```




#Ischemic Heart Disease/100,000:2017
```{r}
IHD<- codes[grepl("[Hh]eart", codes$display), ] 
```

#Deaths per 100,000 of IHD
```{r}
IHD_df <- get_data("SA_0000001444")
```

#2004 is the latest available data here :/
```{r}
IHD_df <- IHD_df %>%
  filter(sex == "Both sexes")
```

```{r}


IHD_df = rename(IHD_df, IHD = value) #rename columns
```

```{r}
Total3 <- merge(Total2, IHD_df, by= "country") #merge mortality index into total df
```

```{r}
Total3 <- subset(Total3, select = c(country, Income_Group, year.x, region.x, Spending_per_capita, GHE_percent, Adult_mort, HIV_AIDS, IHD))    #select desired columns
```



```{r}  
Total3 = rename(Total3, year = year.x)   #rename
Total3= rename(Total3, region= region.x)

```


#Neonatal
#per 1000 live births
#2016
```{r}
Neonatal_Disorders<- codes[grepl("[Nn]eonatal", codes$display), ] 
```

```{r}
Neonatal_Df <- get_data("WHOSIS_000003")
```

```{r}
Neonatal_Df<- Neonatal_Df %>%
  filter(sex == "Both sexes", year== "2016")
```


```{r}


Neonatal_Df = rename(Neonatal_Df, Neonatal = value) #rename columns
```

```{r}
Total4 <- merge(Total3, Neonatal_Df, by= "country") #merge mortality index into total df
```

```{r}
Total4 <- subset(Total4, select = c(country, Income_Group, year.x, region.x, Spending_per_capita, GHE_percent, Adult_mort,HIV_AIDS, IHD, Neonatal))    #select desired columns
```



```{r}  
Total4 = rename(Total4, year = year.x)   #rename
Total4= rename(Total4, region= region.x)

```


#Stroke-- most recent data is from 2004
#per 100,000
```{r}
Stroke<- codes[grepl("[Cc]erebrovascular", codes$display), ] 
```

```{r}
Stroke_Df <- get_data("SA_0000001690")


```

```{r}
Stroke_Df<- Stroke_Df %>%
  filter(sex == "Both sexes")
```

```{r}


Stroke_Df = rename(Stroke_Df, Stroke = value) #rename columns
```

```{r}
Total5 <- merge(Total4, Stroke_Df, by= "country") #merge mortality index into total df
```

```{r}
Total5 <- subset(Total5, select = c(country, Income_Group, year.x, region.x, Spending_per_capita, GHE_percent, Adult_mort, HIV_AIDS, IHD, Neonatal, Stroke))    #select desired columns
```



```{r}  
Total5 = rename(Total5, year = year.x)   #rename
Total5= rename(Total5, region= region.x)

```




#Lower Respiratory Disease-- TB, Pneumonia, Bronchitis (WHO doesnt have info on Pneumonia and Bronch so I just used TB), also TB among non HIV cases because HIV/AIDS is another indicator I will be using so I dont want to double count deaths. Would like to include pneumonia...found data at another source: https://ourworldindata.org/pneumonia
#per 100,000
```{r}
Respiratory<- codes[grepl("tuberculosis", codes$display), ] 
```

```{r}
Respiratory_Df_TB <- get_data("MDG_0000000017")


```

```{r}
Respiratory_Df_TB<- Respiratory_Df_TB %>%
  filter(year== "2016")
```

```{r}


Respiratory_Df_TB = rename(Respiratory_Df_TB, TB_resp = value) #rename columns
```

```{r}
Total6 <- merge(Total5, Respiratory_Df_TB, by= "country") #merge mortality index into total df
```

```{r}
Total6 <- subset(Total6, select = c(country, Income_Group, year.x, region.x, Spending_per_capita, GHE_percent, Adult_mort,HIV_AIDS, IHD, Neonatal, Stroke, TB_resp))    #select desired columns
```



```{r}  
Total6 = rename(Total6, year = year.x)   #rename
Total6= rename(Total6, region= region.x)

```





#Diarrheal Diseases
```{r}
Diarrhoea<- codes[grepl("[Dd]iarrhoea", codes$display), ] 
```

```{r}
Diarrhoea_DF <- get_data("WSH_10")


```

```{r}
Diarrhoea_DF<- Diarrhoea_DF %>%
  filter(year== "2016", sex== "Both sexes", agegroup== "All age groups (total)")
```

```{r}


Diarrhoea_DF = rename(Diarrhoea_DF, Diarrhoea = value) #rename columns
```

```{r}
Total7 <- merge(Total6, Diarrhoea_DF, by= "country") #merge mortality index into total df
```

```{r}
Total7 <- subset(Total7, select = c(country, Income_Group, year.x, region.x, Spending_per_capita, GHE_percent, Adult_mort,HIV_AIDS, IHD, Neonatal, Stroke, TB_resp, Diarrhoea))    #select desired columns
```



```{r}  
Total7 = rename(Total7, year = year.x)   #rename
Total7= rename(Total7, region= region.x)

```


#Road Injuries
```{r}
Road_Injuries<- codes[grepl("[Rr]oad", codes$display), ] 
```

```{r}
Road_Injuries_DF <- get_data("RS_198")


```

```{r}
Road_Injuries_DF<-Road_Injuries_DF %>%
  filter(year== "2016")
```

```{r}


Road_Injuries_DF = rename(Road_Injuries_DF, Road_Injuries = value) #rename columns
```

```{r}
Total8 <- merge(Total7, Road_Injuries_DF, by= "country") #merge mortality index into total df
```

```{r}
Total8 <- subset(Total8, select = c(country, Income_Group, year.x, region.x, Spending_per_capita, GHE_percent, Adult_mort,HIV_AIDS, IHD, Neonatal, Stroke, TB_resp, Diarrhoea, Road_Injuries))    #select desired columns
```



```{r}  
Total8 = rename(Total8, year = year.x)   #rename
Total8= rename(Total8, region= region.x)

```

#Congenital Birth Defects 
#Data is from WHO but not available in the health indiactors API
#https://www.who.int/healthinfo/global_burden_disease/estimates/en/index3.html
#rneo15= number of live births resulting in death from congential birth defects / 1000 live births
#From 2016

```{r}
setwd("~/Documents/")
childhood_estimates <- read.csv("childhood_estimates.csv")
```

```{r}
childhood_estimates <- subset(childhood_estimates, select = c(level, year, rneo15, whoname))    #select desired columns
```

```{r}
childhood_estimates = rename(childhood_estimates, country= whoname)   #rename
childhood_estimates= rename(childhood_estimates, birth_defects = rneo15)
```

```{r}
childhood_estimates<-childhood_estimates %>%
  filter(year== "2016", level=="country")
```

```{r}
Total9 <- merge(Total8, childhood_estimates, by= "country") #merge mortality index into total df
```


```{r}
Total9 <- subset(Total9, select = c(country, Income_Group, year.x, region.x, Spending_per_capita, GHE_percent, Adult_mort,HIV_AIDS, IHD, Neonatal, Stroke, TB_resp, Diarrhoea, Road_Injuries, HIV_AIDS, birth_defects))    #select desired columns
```



```{r}  
Total9 = rename(Total9, year = year.x)   #rename


```



#Malaria
https://apps.who.int/iris/bitstream/handle/10665/259492/9789241565523-eng.pdf?sequence=1
#90% of malaria deaths are from the WHO Africa region. Thus, many countries are not included in this because their malaria deaths are negligable. Most developed countries don't have any malaria deaths per year, and if they do they are often from travel. For this reason, if a first world country is not listed here, I am going to put 0 for malaria deaths, as first world countries in general don't really have any malaria disease burden. 


https://ourworldindata.org/malaria#all-charts-preview
#Deaths per 100,000 individuals
#2016

```{r}
malaria <- read.csv("death-rates-malaria.csv")
```

```{r}
malaria <- subset(malaria, select = c(Malaria, Year, Entity))    #select desired columns
```

```{r}
malaria  = rename(malaria , country= Entity)   #rename
malaria= rename(malaria, year = Year)
```

```{r}
malaria <-malaria  %>%
  filter(year== "2016")
```

```{r}
Total10 <- merge(Total9, malaria, by= "country") #merge mortality index into total df
```


```{r}
Total10 <- subset(Total10, select = c(country, Income_Group, year.x, region.x, Spending_per_capita, GHE_percent, Adult_mort,HIV_AIDS, IHD, Neonatal, Stroke, TB_resp, Diarrhoea, Road_Injuries, HIV_AIDS, birth_defects, Malaria))    #select desired columns
```



```{r}  
Total10 = rename(Total10, year = year.x)   #rename
Total10 = rename(Total10, region= region.x)


```


```{r}
hist5 <-ggplot(data=Total10, aes(Total10$Income_Group)) + 
  geom_histogram(stat="count", fill = "blue") + labs(x= "Income Groups", y = "Count", title = "Distribution of Income Group Counts") + theme(plot.title = element_text(hjust = 0.5))
```

```{r}
hist5
```


#Pneumonia 
#https://ourworldindata.org/pneumonia
```{r}
pn <- read.csv("pneumonia-mortality-by-age.csv")
```

```{r}
pn <- subset(pn, select = c(p1,p2, p3, p4, p5, Year, Entity))    #select desired columns
```

```{r}
pn  = rename(pn , country= Entity)   #rename
pn= rename(pn, year = Year)
```

```{r}
pn<- pn  %>%
  filter(year== "2016")
```

```{r}
pn <- pn %>%
  mutate(pneumonia = (p1+p2+p3+p4+p5))
```



```{r}
Total11 <- merge(Total10, pn, by= "country") #merge mortality index into total df
```


```{r}
Total11 <- subset(Total11, select = c(country, Income_Group, year.x, region, Spending_per_capita, GHE_percent, Adult_mort,HIV_AIDS, IHD, Neonatal, Stroke, TB_resp, Diarrhoea, Road_Injuries, birth_defects, Malaria, pneumonia))    #select desired columns
```



```{r}  
Total11 = rename(Total11, year = year.x)   #rename


```

```{r}
Total11$HIV_AIDS <- as.numeric(gsub("<", "", Total11$HIV_AIDS))

Total11$TB_resp <- as.numeric(gsub("\\[.*?\\]", "", Total11$TB_resp))

Total11$Diarrhoea <- as.numeric(gsub("\\[.*?\\]", "", Total11$Diarrhoea))

```


#Lower Respiratory Diseases

```{r}
Total12 <- Total11 %>%
  mutate(Lower_Respiratory = (TB_resp+ pneumonia))
```


#Put all diseases on a scale of deaths per 100,000
#Neonatal and congenital birth defects are per 1000 live births, so multiply this # by 100

```{r}
Total13 <- Total12 %>%
  mutate(Neonatal_toscale = (Neonatal * 100))
```

```{r}
Total14 <- Total13 %>%
  mutate(Birth_Defects_toscale = birth_defects * 100)
```

#Population

```{r}
population<- codes[grepl("[Pp]opulation", codes$display), ] 
```

```{r}
population_df <- get_data("WHS9_86")


```

```{r}
population_df = rename(population_df , population_1000 = value)   #rename

```

```{r}
population_df<- population_df %>%
  filter(year== "2016")
```




```{r}
Total15 <- merge(Total14, population_df, by= "country") #merge mortality index into total df
```

`
```{r}
Total15 <- subset(Total15, select = c(country, Income_Group, year.x, region.x, Spending_per_capita, GHE_percent, Adult_mort,HIV_AIDS, IHD, Neonatal, Stroke, TB_resp, Diarrhoea, Road_Injuries, birth_defects, Malaria, pneumonia, Lower_Respiratory, Neonatal_toscale, Birth_Defects_toscale, population_1000))    #select desired columns
```

```{r}

Total15$population_1000 <- as.numeric(gsub(" ", "", Total15$population_1000))

Total16 <- Total15 %>%
  mutate(Population = (as.numeric(population_1000) * 1000))
```




```{r}  
Total16 = rename(Total16, year = year.x)   #rename
Total16 = rename(Total16, region= region.x)


```


#Need to get Diarrhea on an 100,000 scale (it is not just the crude # out of total population)

```{r}
Total17 <- Total16 %>%
  mutate(Diarrhoea_toscale = ((Diarrhoea/Population)*100000))
```



#Mutate Disease Burden Index
```{r}
Total18 <- Total17 %>%
  mutate(Disease_Burden_Index = ((IHD*1)+ (Neonatal_toscale*.9) +(Stroke*.8)+ (Lower_Respiratory*.7)+ (Diarrhoea_toscale*.6) + (Road_Injuries*.5) + (HIV_AIDS*.3) + (Birth_Defects_toscale *.2 + Malaria *.1))/9)
```


#Scale index between 0 and 1

```{r}
Total19 <- Total18 %>%
  mutate(Disease_Burden_Index_Scaled = (Disease_Burden_Index- min(Disease_Burden_Index))/((max(Disease_Burden_Index)-min(Disease_Burden_Index))))
```

```{r}
write.csv(Total19, "Total19.csv")
```


```{r}
hist6 <-ggplot(data=Total20, aes(Total20$Income_Group)) + 
  geom_bar(stat="count", fill = "blue") + labs(x= "Income Groups", y = "Count")
```

```{r}
hist6
```

```{r}

hist7 <-ggplot(data=Total20, aes(Total20$region)) + 
  geom_bar(stat="count", fill = "red") + labs(x= "Regions", y = "Count")

```


```{r}
hist7
```


```{r}
plot1 <- ggplot(Total20,aes(Spending_per_capita_log, Adult_mort)) + geom_point(aes(col=Income_Group)) + labs(y= "Adult Mortality Rate per 1000 people", x= "Government Healthcare Spending per Capita") + geom_smooth(method= "lm")
```

```{r}
plot1
```

```{r}
lm1 <- lm(Adult_mort ~ Spending_per_capita_log, data=Total20) 
summary(lm1)
```


```{r}
Total20 <- Total19 %>%
  mutate(Spending_per_capita_log = log(Spending_per_capita))
```


#Put an Lm on this 

```{r}
plot2 <- ggplot(Total20,aes( Spending_per_capita_log, Disease_Burden_Index_Scaled)) + geom_point(aes(col=Income_Group)) + labs(y= "Disease Burden Index", x= "Log Government Healthcare Spending per Capita (USD)") + geom_smooth(method = "lm")
```

```{r}
plot2
```


```{r}
lm3 <- lm(Disease_Burden_Index_Scaled ~ GHE_percent, data=Total20) 
summary(lm3)
```

```{r}
plot5 <- ggplot(Total20, aes(Total20$Disease_Burden_Index_Scaled)) +geom_histogram(bins=10, fill="blue") +labs(x= "Disease Burden Index")
plot5
```


```{r}
plot5 <-ggplot(data=Total20, aes(Income_Group, Spending_per_capita)) + 
  geom_bar(stat="summary", fill = "red") + labs(x= "Income Groups", y = "Healthacre Spending per Capita (USD)")
plot5
```


```{r}
plot6 <-ggplot(data=Total20, aes(region, Spending_per_capita)) + 
  geom_bar(stat="summary", fill = "red") + labs(x= "Regions", y = "Healthacre Spending per Capita (USD)")
plot6
```

```{r}
plot7 <-ggplot(data=Total20, aes(region, Disease_Burden_Index_Scaled)) + 
  geom_bar(stat="summary", fill = "purple") + labs(x= "Regions", y = "Healthacre Spending per Capita (USD)")
plot7
```




```{r}
plot8 <-ggplot(data=Total20, aes(Income_Group, Disease_Burden_Index_Scaled)) + 
  geom_bar(stat="summary", fill = "purple") + labs(x= "Regions", y = "Healthacre Spending per Capita (USD)")
plot8
```


#Filtering by income group and region

```{r}
Total21 <- Total20 %>%
  filter(Income_Group== "High-income")
```

```{r}
plot6 <- ggplot(Total21,aes( Spending_per_capita_log, Disease_Burden_Index_Scaled)) + geom_point(aes(col=region)) + labs(y= "Disease Burden Index", x= "Log Government Healthcare Spending per Capita (USD)") + geom_smooth(method = "lm")
plot6
```

```{r}
lm4 <- lm(Disease_Burden_Index_Scaled ~ Spending_per_capita_log + region, data=Total21) 
summary(lm4)
```


```{r}
Total22 <- Total20 %>%
  filter(Income_Group== "Lower-middle-income")
```

```{r}
plot6 <- ggplot(Total22,aes( Spending_per_capita_log, Disease_Burden_Index_Scaled)) + geom_point(aes(col=region)) + labs(y= "Disease Burden Index", x= "Log Government Healthcare Spending per Capita (USD)") + geom_smooth(method = "lm")
plot6
```

```{r}
lm5 <- lm(Disease_Burden_Index_Scaled ~ Spending_per_capita_log + region, data=Total22) 
summary(lm5)
```

```{r}
Total23 <- Total20 %>%
  filter(Income_Group== "Low-income")
```

```{r}
plot6 <- ggplot(Total23,aes( Spending_per_capita_log, Disease_Burden_Index_Scaled)) + geom_point(aes(col=region)) + labs(y= "Disease Burden Index", x= "Log Government Healthcare Spending per Capita (USD)") + geom_smooth(method = "lm")
plot6
```

```{r}
lm6 <- lm(Disease_Burden_Index_Scaled ~ Spending_per_capita_log + region, data=Total23) 
summary(lm6)
```

```{r}
Total24 <- Total20 %>%
  filter(Income_Group== "Upper-middle-income")
```

```{r}
plot6 <- ggplot(Total24,aes( Spending_per_capita_log, Disease_Burden_Index_Scaled)) + geom_point(aes(col=region)) + labs(y= "Disease Burden Index", x= "Log Government Healthcare Spending per Capita (USD)") + geom_smooth(method = "lm")
plot6
```


```{r}
lm7 <- lm(Disease_Burden_Index_Scaled ~ Spending_per_capita_log + region, data=Total24) 
summary(lm7)
```



```{r}
Total25 <- Total20 %>%
  filter(region == "Americas")
```



```{r}
plot7 <- ggplot(Total25,aes( Spending_per_capita_log, Disease_Burden_Index_Scaled)) + geom_point(aes(col=Income_Group)) + labs(y= "Disease Burden Index", x= "Log Government Healthcare Spending per Capita (USD)") + geom_smooth(method = "lm")
plot7
```


```{r}
lm8 <- lm(Disease_Burden_Index_Scaled ~ Spending_per_capita_log + Income_Group, data=Total25) 
summary(lm8)
```

```{r}
Total26 <- Total20 %>%
  filter(region == "Europe")
```



```{r}
plot7 <- ggplot(Total26,aes( Spending_per_capita_log, Disease_Burden_Index_Scaled)) + geom_point(aes(col=Income_Group)) + labs(y= "Disease Burden Index", x= "Log Government Healthcare Spending per Capita (USD)") + geom_smooth(method = "lm")
plot7
```



```{r}
lm8 <- lm(Disease_Burden_Index_Scaled ~ Spending_per_capita_log + Income_Group, data=Total26) 
summary(lm8)
```


```{r}
Total27 <- Total20 %>%
  filter(region == "Africa")
```



```{r}
plot7 <- ggplot(Total27,aes( Spending_per_capita_log, Disease_Burden_Index_Scaled)) + geom_point(aes(col=Income_Group)) + labs(y= "Disease Burden Index", x= "Log Government Healthcare Spending per Capita (USD)") + geom_smooth(method = "lm")
plot7
```


```{r}
lm8 <- lm(Disease_Burden_Index_Scaled ~ Spending_per_capita_log + Income_Group, data=Total27) 
summary(lm8)
```

#Only 7 observations in eastern med so ...not legit
#Rule of thumb: if there are over 10 observations then I will do the regression
```{r}
Total28 <- Total20 %>%
  filter(region == "Eastern Mediterranean")
```


#Western Pacific...only 10 observations so kind of iffy
```{r}
Total29 <- Total20 %>%
  filter(region == "Western Pacific")
```

```{r}
plot7 <- ggplot(Total29,aes( Spending_per_capita_log, Disease_Burden_Index_Scaled)) + geom_point(aes(col=Income_Group)) + labs(y= "Disease Burden Index", x= "Log Government Healthcare Spending per Capita (USD)") + geom_smooth(method = "lm")
plot7
```

```{r}
lm9 <- lm(Disease_Burden_Index_Scaled ~ Spending_per_capita_log + Income_Group, data=Total29) 
summary(lm9)
```

```{r}
Total30 <- Total20 %>%
  filter(region == "South-East Asia")
```

```{r}
lm9 <- lm(Disease_Burden_Index_Scaled ~ Spending_per_capita_log + Income_Group, data=Total30) 
summary(lm9)
```


#Searching for indactors of healthcare systems (private vs publics sector or universal)


```{r}
national<- codes[grepl("UHC", codes$display), ] 
```

```{r}
national_df <- get_data("HSS_UHCLEGISLATION")


```



```{r}
national_df = rename(national_df, Universal_Legislation = value) #rename columns
```

```{r}
Total31<- merge(Total20 , national_df, by= "country") #merge mortality index into total df
```


```{r}
Total31 <- subset(Total31, select = c(country, Income_Group, year.x, region.x, Spending_per_capita, GHE_percent, Adult_mort,HIV_AIDS, IHD, Neonatal, Stroke, TB_resp, Diarrhoea, Road_Injuries, birth_defects, Malaria, pneumonia, Lower_Respiratory, Neonatal_toscale, Birth_Defects_toscale, population_1000, Disease_Burden_Index_Scaled, Spending_per_capita_log, Universal_Legislation))    #select desired columns
```



```{r}  
Total31 = rename(Total31, year = year.x)   #rename
Total31 = rename(Total31, region = region.x) 


```




```{r}
plot6 <- ggplot(Total31,aes( Spending_per_capita_log, Disease_Burden_Index_Scaled)) + geom_point(aes(col=Universal_Legislation)) + labs(y= "Disease Burden Index", x= "Log Government Healthcare Spending per Capita (USD)") + geom_smooth(method = "lm")
plot6
```

```{r}
lm9 <- lm(Disease_Burden_Index_Scaled ~ Spending_per_capita_log+ region , data=Total31) 
summary(lm9)
```

```{r}
plot5 <-ggplot(data=Total31, aes(Universal_Legislation, Spending_per_capita)) + 
  geom_bar(stat="summary", fill = "red") + labs(x= "Passed Universal Healthcare Legislation", y = "Healthacre Spending per Capita (USD)")
plot5
```





```{r}
plot5 <-ggplot(data=Total31, aes(Universal_Legislation, Disease_Burden_Index_Scaled)) + 
  geom_bar(stat="summary", fill = "red") + labs(x= "Passed Universal Healthcare Legislation", y = "Disease Burden Index")
plot5
```


```{r}
plot5 <-ggplot(data=Total31, aes(Universal_Legislation, GHE_percent)) + 
  geom_bar(stat="summary", fill = "red") + labs(x= "Passed Universal Healthcare Legislation", y = "Government Healthcare Spending as a Percent of GDP")
plot5
```










```{r}
install.packages("Hmisc")
library("Hmisc")
```


```{r}
Total21 <- subset(Total20, select = c(, Income_Group, year.x, region, Spending_per_capita, GHE_percent, Adult_mort,HIV_AIDS, IHD, Neonatal, Stroke, TB_resp, Diarrhoea, Road_Injuries, birth_defects, Malaria, pneumonia))    #select desired columns
```



```{r}
Total20.cor = cor(Total20)
```


```{r}
Total20.rcorr = rcorr(as.matrix(Total20))
mydata.rcorr
```




```{r}
install.packages("corrplot")
library(corrplot)
```

```{r}
library(corrplot)
corrplot(Total20.cor)
```



library(ggplot2)
library(ggcorrplot)

# Correlation matrix
data(mtcars)
corr <- round(cor(mtcars), 1)

# Plot
ggcorrplot(corr, hc.order = TRUE, 
           type = "lower", 
           lab = TRUE, 
           lab_size = 3, 
           method="circle", 
           colors = c("tomato2", "white", "springgreen3"), 
           title="Correlogram of mtcars", 
           ggtheme=theme_bw)






