---
title: "WHO_DA 401 Project.Rmd"
author: "Maddy Matura"
date: "February 29, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Installing packages
```{r}
#Install WHO package created by expersso
install.packages(devtools)
library(devtools)
install_github("expersso/WHO")
```

```{r}
library(WHO)
library(dplyr)
library(tidyr)
library(ggplot2)


```


```{r}
#Look at indicators with the word expenditures
expenditures <- codes[grepl("expend", codes$display), ]
```

```{r}
#look at indicators with the word GDP
GDP <- codes[grepl("GDP", codes$display), ]
```

```{r}
#Create GDP dataframe
GDP_df <- get_data("GHED_CHEGDP_SHA2011")
```

```{r}
GHE<- codes[grepl("GHE", codes$display), ] 
```

#get data using API & WHO package
```{r}
GHE_percent_GDP <- get_data("GHED_GGHE-DGDP_SHA2011")
```

#get spending per capita data using API & WHO package
```{r}
Spending_per_capita_USD<- get_data("GHED_GGHE-D_pc_US_SHA2011") #Spending per capita in USD for each country
```

#Clean GHE data
```{r}
GHE_percent_GDP16 <- GHE_percent_GDP %>%
  filter(year == "2016")  #filter for 2016
GHE_percent_GDP16 <- subset(GHE_percent_GDP16, select = c(worldbankincomegroup, year,region, country, value))  #select desired columns (income group, year, region, country, value)
GHE_percent_GDP16_noNA <- GHE_percent_GDP16 %>% drop_na()  #get rid of NA
GHE_percent_GDP16_noNA = rename(GHE_percent_GDP16_noNA, GHE_percent = value) #rename columns
```


#Clean spending per capita data
```{r}

Spending_per_capita_USD16 <- Spending_per_capita_USD %>%
  filter(year == "2016")  #filter for 2016 only
Spending_per_capita_USD16<- subset(Spending_per_capita_USD16, select = c(worldbankincomegroup, year,region, country, value)) #subset with desired columns

Spending_per_capita_USD16_noNA <- Spending_per_capita_USD16 %>% drop_na() #drop na

Spending_per_capita_USD16_noNA = rename(Spending_per_capita_USD16_noNA, Spending_per_capita = value)  #rename columns
```


#Merge spending and GHE data by country
```{r}
Total <- merge(Spending_per_capita_USD16_noNA, GHE_percent_GDP16_noNA, by= "country") #merge dataframes to get total
```

```{r}
summarize(Total)
```


#Subset desired columns
```{r}
Total <- subset(Total, select = c(country, worldbankincomegroup.x, year.x, region.x, Spending_per_capita, GHE_percent)) #pull out desired columns

```


```{r}

Total = rename(Total, worldbankincomegroup = worldbankincomegroup.x)  #rename all columns
Total = rename(Total, year = year.x)
Total= rename(Total, region= region.x)
```



#Mortality: Adult mortality per 1000 people in between ages 15-60
```{r}
Mort <- codes[grepl("[Mm]ortality", codes$display), ]   #search for mortality health indicator
```

```{r}
Mort_df <- get_data("WHOSIS_000004")   #make the mortality indicator into df
```

```{r}
Adult_Mort_2016 <- Mort_df %>% 
  filter(year == "2016", sex== "Both sexes") #filter by 2016 and both sexes


```

```{r}

Adult_Mort_2016_noNA <- Adult_Mort_2016 %>% drop_na()  #drop na

Adult_Mort_2016_noNA = rename(Adult_Mort_2016_noNA, Adult_mort = value) #rename columns
```

```{r}
Total <- merge(Total, Adult_Mort_2016_noNA, by= "country") #merge mortality index into total df
```

```{r}
Total <- subset(Total, select = c(country, worldbankincomegroup, year.x, region.x, Spending_per_capita, GHE_percent, Adult_mort))    #select desired columns
```


```{r}  
Total = rename(Total, year = year.x)   #rename
Total= rename(Total, region= region.x)

```

```{r}
Total = rename(Total, Income_Group = worldbankincomegroup) 
```



#Disease Burden Index Calculation
#Over all ages and all genders

#HIV/AIDS
#Deaths per 100,000. The year 2012 is being used because this is the most recent data available
#HIV needs to be merged to the dataset first or else the country count will be too low if added after the rest of the health indicators
```{r}
HIV<- codes[grepl("HIV", codes$display), ] #search for HIV in indicators
```

```{r}
HIV_DF <- get_data("WHS2_138")   #get the data using the API, turn into dataframe


```

```{r}
HIV_DF<-HIV_DF %>%
  filter(year== "2012")    #filter for year 2012
```

```{r}


HIV_DF = rename(HIV_DF, HIV_AIDS = value) #rename columns
```

```{r}
Total2 <- merge(Total, HIV_DF, by= "country") #merge mortality index into total df
```

```{r}
Total2 <- subset(Total2, select = c(country, Income_Group, year.x, region.x, Spending_per_capita, GHE_percent, Adult_mort, HIV_AIDS))    #select desired columns
```


```{r}  
Total2 = rename(Total2, year = year.x)   #rename
Total2= rename(Total2, region= region.x)

```


#Ischemic Heart Disease deaths per 100,000, from year 2004
```{r}
IHD<- codes[grepl("[Hh]eart", codes$display), ] #search for heart disease
```


```{r}
IHD_df <- get_data("SA_0000001444")
```


```{r}
IHD_df <- IHD_df %>%
  filter(sex == "Both sexes")
```

```{r}


IHD_df = rename(IHD_df, IHD = value) #rename columns
```

```{r}
Total3 <- merge(Total2, IHD_df, by= "country") #merge mortality index into total df
```

```{r}
Total3 <- subset(Total3, select = c(country, Income_Group, year.x, region.x, Spending_per_capita, GHE_percent, Adult_mort, HIV_AIDS, IHD))    #select desired columns
```


```{r}  
Total3 = rename(Total3, year = year.x)   #rename
Total3= rename(Total3, region= region.x)

```


#Neonatal deaths per 1000 live births in 2016
```{r}
Neonatal_Disorders<- codes[grepl("[Nn]eonatal", codes$display), ]  #search for neonatal in health indicators
```

```{r}
Neonatal_Df <- get_data("WHOSIS_000003")
```

```{r}
Neonatal_Df<- Neonatal_Df %>%
  filter(sex == "Both sexes", year== "2016")    #filter for both sexes and 2016
```


```{r}
Neonatal_Df = rename(Neonatal_Df, Neonatal = value) #rename columns
```

```{r}
Total4 <- merge(Total3, Neonatal_Df, by= "country") #merge mortality index into total df
```

```{r}
Total4 <- subset(Total4, select = c(country, Income_Group, year.x, region.x, Spending_per_capita, GHE_percent, Adult_mort,HIV_AIDS, IHD, Neonatal))    #select desired columns
```



```{r}  
Total4 = rename(Total4, year = year.x)   #rename
Total4= rename(Total4, region= region.x)

```


#Stroke deaths per 100,000 in 2004
#2004 is the most recent available data
```{r}
Stroke<- codes[grepl("[Cc]erebrovascular", codes$display), ] #search for cerebrovascular, this is the medical term for a stroke
```

```{r}
Stroke_Df <- get_data("SA_0000001690")


```

```{r}
Stroke_Df<- Stroke_Df %>%
  filter(sex == "Both sexes")
```

```{r}


Stroke_Df = rename(Stroke_Df, Stroke = value) #rename columns
```

```{r}
Total5 <- merge(Total4, Stroke_Df, by= "country") #merge mortality index into total df
```

```{r}
Total5 <- subset(Total5, select = c(country, Income_Group, year.x, region.x, Spending_per_capita, GHE_percent, Adult_mort, HIV_AIDS, IHD, Neonatal, Stroke))    #select desired columns
```



```{r}  
Total5 = rename(Total5, year = year.x)   #rename
Total5= rename(Total5, region= region.x)

```




#Lower Respiratory Disease
#Lower respiratory disease can include tuberculosis, pneumonia, and bronchitis
#The WHO only surveys tuberculosis, so this was searched for
#Later, a dataset on pneumonia from an alternate source will be combined with the TB dataset: https://ourworldindata.org/pneumonia
#I could not find a dataset with numbers on bronchitis

#Non HIV TB deaths/ 100,000 in 2016
```{r}
Respiratory<- codes[grepl("tuberculosis", codes$display), ] 
```

```{r}
Respiratory_Df_TB <- get_data("MDG_0000000017")


```

```{r}
Respiratory_Df_TB<- Respiratory_Df_TB %>%
  filter(year== "2016")
```

```{r}


Respiratory_Df_TB = rename(Respiratory_Df_TB, TB_resp = value) #rename columns
```

```{r}
Total6 <- merge(Total5, Respiratory_Df_TB, by= "country") #merge mortality index into total df
```

```{r}
Total6 <- subset(Total6, select = c(country, Income_Group, year.x, region.x, Spending_per_capita, GHE_percent, Adult_mort,HIV_AIDS, IHD, Neonatal, Stroke, TB_resp))    #select desired columns
```

```{r}  
Total6 = rename(Total6, year = year.x)   #rename
Total6= rename(Total6, region= region.x)

```


#Diarrheal deaths per the entire population in 2016
#This indicator is not out of 100,000 so it will later have to be divided by the population to make it out of 100,000
```{r}
Diarrhoea<- codes[grepl("[Dd]iarrhoea", codes$display), ] 
```

```{r}
Diarrhoea_DF <- get_data("WSH_10")


```

```{r}
Diarrhoea_DF<- Diarrhoea_DF %>%
  filter(year== "2016", sex== "Both sexes", agegroup== "All age groups (total)")
```

```{r}
Diarrhoea_DF = rename(Diarrhoea_DF, Diarrhoea = value) #rename columns
```

```{r}
Total7 <- merge(Total6, Diarrhoea_DF, by= "country") #merge mortality index into total df
```

```{r}
Total7 <- subset(Total7, select = c(country, Income_Group, year.x, region.x, Spending_per_capita, GHE_percent, Adult_mort,HIV_AIDS, IHD, Neonatal, Stroke, TB_resp, Diarrhoea))    #select desired columns
```

```{r}  
Total7 = rename(Total7, year = year.x)   #rename
Total7= rename(Total7, region= region.x)

```


#Road Injury deaths per 100,000 in 2016
```{r}
Road_Injuries<- codes[grepl("[Rr]oad", codes$display), ] 
```

```{r}
Road_Injuries_DF <- get_data("RS_198")


```

```{r}
Road_Injuries_DF<-Road_Injuries_DF %>%
  filter(year== "2016")
```

```{r}


Road_Injuries_DF = rename(Road_Injuries_DF, Road_Injuries = value) #rename columns
```

```{r}
Total8 <- merge(Total7, Road_Injuries_DF, by= "country") #merge mortality index into total df
```

```{r}
Total8 <- subset(Total8, select = c(country, Income_Group, year.x, region.x, Spending_per_capita, GHE_percent, Adult_mort,HIV_AIDS, IHD, Neonatal, Stroke, TB_resp, Diarrhoea, Road_Injuries))    #select desired columns
```


```{r}  
Total8 = rename(Total8, year = year.x)   #rename
Total8= rename(Total8, region= region.x)

```

#Congenital Birth Defects 
#Data is from WHO but not available in the health indiactors API
#https://www.who.int/healthinfo/global_burden_disease/estimates/en/index3.html
#rneo15= number of live births resulting in death from congential birth defects / 1000 live births
#From 2016
```{r}
setwd("~/Documents/")
childhood_estimates <- read.csv("childhood_estimates.csv")  #import dataset 
```

```{r}
childhood_estimates <- subset(childhood_estimates, select = c(level, year, rneo15, whoname))    #select desired columns
```

```{r}
childhood_estimates = rename(childhood_estimates, country= whoname)   #rename
childhood_estimates= rename(childhood_estimates, birth_defects = rneo15)
```

```{r}
childhood_estimates<-childhood_estimates %>%
  filter(year== "2016", level=="country")
```

```{r}
Total9 <- merge(Total8, childhood_estimates, by= "country") #merge mortality index into total df
```

```{r}
Total9 <- subset(Total9, select = c(country, Income_Group, year.x, region.x, Spending_per_capita, GHE_percent, Adult_mort,HIV_AIDS, IHD, Neonatal, Stroke, TB_resp, Diarrhoea, Road_Injuries, HIV_AIDS, birth_defects))    #select desired columns
```

```{r}  
Total9 = rename(Total9, year = year.x)   #rename


```



#Malaria
#https://apps.who.int/iris/bitstream/handle/10665/259492/9789241565523-eng.pdf?sequence=1
#90% of malaria deaths are from the WHO Africa region. Thus, many countries are not included in this because their malaria deaths are negligable. Most developed countries don't have any malaria deaths per year, and if they do they are often from travel.  

#I have found an alternate dataset because the WHO does not record malaria deaths in high income countries
#https://ourworldindata.org/malaria#all-charts-preview
#Deaths per 100,000 individuals
#2016

```{r}
malaria <- read.csv("death-rates-malaria.csv") #import dataset
```

```{r}
malaria <- subset(malaria, select = c(Malaria, Year, Entity))    #select desired columns
```

```{r}
malaria  = rename(malaria , country= Entity)   #rename
malaria= rename(malaria, year = Year)
```

```{r}
malaria <-malaria  %>%
  filter(year== "2016")
```

```{r}
Total10 <- merge(Total9, malaria, by= "country") #merge mortality index into total df
```

```{r}
Total10 <- subset(Total10, select = c(country, Income_Group, year.x, region.x, Spending_per_capita, GHE_percent, Adult_mort,HIV_AIDS, IHD, Neonatal, Stroke, TB_resp, Diarrhoea, Road_Injuries, HIV_AIDS, birth_defects, Malaria))    #select desired columns
```

```{r}  
Total10 = rename(Total10, year = year.x)   #rename
Total10 = rename(Total10, region= region.x)


```


#Pneumonia 
#https://ourworldindata.org/pneumonia
#These numbers are death per 100,000 people in 2016
#Will be added later to TB deaths in order to account for lower respiratory deaths
```{r}
pn <- read.csv("pneumonia-mortality-by-age.csv")  #import dataset
```

```{r}
pn <- subset(pn, select = c(p1,p2, p3, p4, p5, Year, Entity))    #select desired columns
```

```{r}
pn  = rename(pn , country= Entity)   #rename
pn= rename(pn, year = Year)
```

```{r}
pn<- pn  %>%
  filter(year== "2016")
```

```{r}
pn <- pn %>%
  mutate(pneumonia = (p1+p2+p3+p4+p5))
```

```{r}
Total11 <- merge(Total10, pn, by= "country") #merge mortality index into total df
```

```{r}
Total11 <- subset(Total11, select = c(country, Income_Group, year.x, region, Spending_per_capita, GHE_percent, Adult_mort,HIV_AIDS, IHD, Neonatal, Stroke, TB_resp, Diarrhoea, Road_Injuries, birth_defects, Malaria, pneumonia))    #select desired columns
```

```{r}  
Total11 = rename(Total11, year = year.x)   #rename


```




#Clean dataset so that disease burden index can be calculated
```{r}
Total11$HIV_AIDS <- as.numeric(gsub("<", "", Total11$HIV_AIDS))  #get rid of < and replace with nothing

Total11$TB_resp <- as.numeric(gsub("\\[.*?\\]", "", Total11$TB_resp))  #get rid of [#-#] and replace with nothing

Total11$Diarrhoea <- as.numeric(gsub("\\[.*?\\]", "", Total11$Diarrhoea))  

```


#Lower Respiratory Diseases
#mutate a new column to include Tb and pneumonia deaths per 100,000

```{r}
Total12 <- Total11 %>%
  mutate(Lower_Respiratory = (TB_resp+ pneumonia))
```


#Put all diseases on a scale of deaths per 100,000
#Neonatal and congenital birth defects are per 1000 live births, so multiply this # by 100

```{r}
Total13 <- Total12 %>%
  mutate(Neonatal_toscale = (Neonatal * 100))
```

```{r}
Total14 <- Total13 %>%
  mutate(Birth_Defects_toscale = birth_defects * 100)
```

#Population
#Put population column in dataset so that diarreheal disease can be mutated to per 100,000 people
```{r}
population<- codes[grepl("[Pp]opulation", codes$display), ] 
```

```{r}
population_df <- get_data("WHS9_86")


```

```{r}
population_df = rename(population_df , population_1000 = value)   #rename

```

```{r}
population_df<- population_df %>%  #filter for populations in 2016
  filter(year== "2016")
```

```{r}
Total15 <- merge(Total14, population_df, by= "country") #merge mortality index into total df
```

```{r}
Total15 <- subset(Total15, select = c(country, Income_Group, year.x, region.x, Spending_per_capita, GHE_percent, Adult_mort,HIV_AIDS, IHD, Neonatal, Stroke, TB_resp, Diarrhoea, Road_Injuries, birth_defects, Malaria, pneumonia, Lower_Respiratory, Neonatal_toscale, Birth_Defects_toscale, population_1000))    #select desired columns
```


#Population is written in the 1000's, so needs to be multipled by 1000 to get the total population of each country
```{r}

Total15$population_1000 <- as.numeric(gsub(" ", "", Total15$population_1000))

Total16 <- Total15 %>%
  mutate(Population = (as.numeric(population_1000) * 1000))
```

```{r}  
Total16 = rename(Total16, year = year.x)   #rename
Total16 = rename(Total16, region= region.x)


```


#Need to get Diarrhea on the deaths per 100,000 scale 
```{r}
Total17 <- Total16 %>%
  mutate(Diarrhoea_toscale = ((Diarrhoea/Population)*100000))
```



#Mutate Disease Burden Index
#Add up all indicators for a country and weight by number of the top 10 causes of early death list
```{r}
Total18 <- Total17 %>%
  mutate(Disease_Burden_Index = ((IHD*1)+ (Neonatal_toscale*.9) +(Stroke*.8)+ (Lower_Respiratory*.7)+ (Diarrhoea_toscale*.6) + (Road_Injuries*.5) + (HIV_AIDS*.3) + (Birth_Defects_toscale *.2 + Malaria *.1))/9)
```


#Scale index between 0 and 1 via min max scaling
```{r}
Total19 <- Total18 %>%
  mutate(Disease_Burden_Index_Scaled = (Disease_Burden_Index- min(Disease_Burden_Index))/((max(Disease_Burden_Index)-min(Disease_Burden_Index))))
```

#Mutate spending per capita to log version, Total 20 is the final dataset
```{r}
Total20 <- Total19 %>%
  mutate(Spending_per_capita_log = log(Spending_per_capita))
```


#Export final data set
```{r}
write.csv(Total20, "Final_Data.csv")
```


#Group dataset by income group and count
```{r}
TotalB <- Total20 %>%
  group_by(Income_Group)%>%
  tally()
```

#Plot income group by counts
```{r}
plot6 <-ggplot(data=TotalB, aes(x= reorder(Income_Group, n), y= n)) + 
  geom_bar(stat="summary", fill = "blue") + labs(x= "Income Groups", y = "Count")
plot6
```



#Filter Western pacific region into separate dataset
```{r}
Total35 <- Total20 %>%
  filter(region == "Western Pacific")
```


#Count by region
```{r}
TotalA <- Total20 %>%
  group_by(region)%>%
  tally()
```


#Plot region counts on bar chart
```{r}
plot5 <-ggplot(data=TotalA, aes(x= reorder(region, n), y= n)) + 
  geom_bar(stat="summary", fill = "red") + labs(x= "Regions", y = "Count")
plot5
```

```{r}
library(gridExtra)
```


```{r}
grid.arrange(plot6, plot5, ncol = 1)
```



#Linear regression of log GH spending and adult mortality
#Colored by region
```{r}
plot1 <- ggplot(Total20,aes(Spending_per_capita_log, Adult_mort)) + geom_point(aes(col=Income_Group)) + labs(y= "Adult Mortality Rate per 1000 people", x= "Log Government Healthcare Spending per Capita (USD)") + geom_smooth(method= "lm")
```



```{r}
plot1
```

```{r}
lm1 <- lm(Adult_mort ~ Spending_per_capita_log, data=Total20) 
summary(lm1)
```


#Validation
```{r}


lm2<- cbind(Total20,lm1$residuals, lm1$fitted.values)
lm2 <- lm2 %>%
  rename(res = "lm1$residuals", fit= "lm1$fitted.values")   

lm2%>%
  ggplot(aes(res)) +
  geom_histogram(bins=15, color="blue", fill="light blue")  #residual histogram

lm2%>%
  ggplot(aes(fit, res)) +       #residual vs fitted plot
  geom_point() 

shapiro.test(lm2$res)  #shapiro test
qqnorm(lm2$res)        #QQ norm plot


```


#Linear regresion of adult mortality by spendering per capita log
#Colored by region
```{r}
plot2 <- ggplot(Total20,aes( Spending_per_capita_log, Disease_Burden_Index_Scaled)) + geom_point(aes(col=region)) + labs(y= "Disease Burden Index", x= "Log Government Healthcare Spending per Capita (USD)") + geom_smooth(method = "lm")
```

```{r}
plot2
```


```{r}
lm1 <- lm(Disease_Burden_Index_Scaled ~ Spending_per_capita_log, data=Total20) 
summary(lm1)
```


#Validation
```{r}


lm2<- cbind(Total20,lm1$residuals, lm1$fitted.values)
lm2 <- lm2 %>%
  rename(res = "lm1$residuals", fit= "lm1$fitted.values")

lm2%>%
  ggplot(aes(res)) +
  geom_histogram(bins=15, color="blue", fill="light blue")   #residual histogram

lm2%>%
  ggplot(aes(fit, res)) +   #res vs fitfed values plot
  geom_point() 

shapiro.test(lm2$res)   #shaprio test
qqnorm(lm2$res)         #QQ norm plot


```


#Linear model with Spending, Income Group, and region predicting Disease burden
```{r}
lm1 <- lm(Disease_Burden_Index_Scaled ~ Spending_per_capita_log +Income_Group + region, data=Total20) 
summary(lm1)
```

#Validation
```{r}


lm2<- cbind(Total20,lm1$residuals, lm1$fitted.values)
lm2 <- lm2 %>%
  rename(res = "lm1$residuals", fit= "lm1$fitted.values")

lm2%>%
  ggplot(aes(res)) +
  geom_histogram(bins=15, color="blue", fill="light blue")

lm2%>%
  ggplot(aes(fit, res)) +
  geom_point() 

shapiro.test(lm2$res)
qqnorm(lm2$res)


```


#Linear model with spending, income group, and region predicing adult mortality
```{r}
lm1 <- lm(Adult_mort ~ Spending_per_capita_log +Income_Group + region, data=Total20) 
summary(lm1)
```


#Validation
```{r}


lm2<- cbind(Total20,lm1$residuals, lm1$fitted.values)
lm2 <- lm2 %>%
  rename(res = "lm1$residuals", fit= "lm1$fitted.values")

lm2%>%
  ggplot(aes(res)) +
  geom_histogram(bins=15, color="blue", fill="light blue")

lm2%>%
  ggplot(aes(fit, res)) +
  geom_point() 

shapiro.test(lm2$res)
qqnorm(lm2$res)


```


#Histogram of log spending distribution
```{r}
plot5 <- ggplot(Total20, aes(Total20$Spending_per_capita_log)) +geom_histogram(bins=7, fill="red") +labs(x= "Log Healthcare Spending per Capita (USD)")
plot5
```


#Histogram of Disease Burden Index Distribution
```{r}
plot5 <- ggplot(Total20, aes(Total20$Disease_Burden_Index_Scaled)) +geom_histogram(bins=10, fill="blue") +labs(x= "Disease Burden Index")
plot5
```

```{r}
library(ggplot2)
```


#Average healthcare spending per income group
```{r}
plot5 <-ggplot(data=Total20, aes(x= reorder(Income_Group, Spending_per_capita), y= Spending_per_capita)) + 
  geom_bar(stat="summary", fill = "blue") + labs(x= "Income Groups", y = "Healthacre Spending per Capita (USD)")
plot5
```

```{r}
grid.arrange(p1, plot5, nrow = 1)
```



#Average spending per income group
```{r}
Sum <- Total20 %>%
  group_by(Income_Group)%>% summarise(mean1= mean(Spending_per_capita))
```

```{r}
Sum <- Total20 %>%
  group_by(region)%>% summarise(mean1= mean(Spending_per_capita))
```



#Average spending per region
```{r}
plot6 <-ggplot(data=Total20, aes(x=reorder(region,Spending_per_capita), Spending_per_capita)) + 
  geom_bar(stat="summary", fill = "red") + labs(x= "Regions", y = "Healthacre Spending per Capita (USD)")
plot6
```


#Average disease burden per region
```{r}
plot7 <-ggplot(data=Total20, aes(x=reorder(region, Disease_Burden_Index_Scaled), Disease_Burden_Index_Scaled)) + 
  geom_bar(stat="summary", fill = "blue") + labs(x= "Regions", y = "Disease Burden Index")
plot7
```



#Average Disease burden per income group
```{r}
plot8 <-ggplot(data=Total20, aes(x=reorder(Income_Group,-Disease_Burden_Index_Scaled), Disease_Burden_Index_Scaled)) + 
  geom_bar(stat="summary", fill = "blue") + labs(x= "Income Groups", y = "Disease Burden Index")
plot8
```


#Filtering by income group and region

```{r}
Total21 <- Total20 %>%
  filter(Income_Group== "High-income")
```

```{r}
plot6 <- ggplot(Total21,aes( Spending_per_capita_log, Disease_Burden_Index_Scaled)) + geom_point() + labs(y= "Disease Burden Index", x= "Log GHS High Income") + geom_smooth(method = "lm")
plot6
```

#Linear model of DBI by spending for High income
```{r}
lm1 <- lm(Disease_Burden_Index_Scaled ~ Spending_per_capita_log, data=Total21) 
summary(lm1)
```


#Validation
```{r}


lm2<- cbind(Total21,lm1$residuals, lm1$fitted.values)
lm2 <- lm2 %>%
  rename(res = "lm1$residuals", fit= "lm1$fitted.values")

lm2%>%
  ggplot(aes(res)) +
  geom_histogram(bins=15, color="blue", fill="light blue")

lm2%>%
  ggplot(aes(fit, res)) +
  geom_point() 

shapiro.test(lm2$res)
qqnorm(lm2$res)


```






#Linear model for lower middle income group
```{r}
Total22 <- Total20 %>%
  filter(Income_Group== "Lower-middle-income")  #filter for LMI
```

```{r}
plot7 <- ggplot(Total22,aes( Spending_per_capita_log, Disease_Burden_Index_Scaled)) + geom_point() + labs(y= "Disease Burden Index", x= "Log GHS Lower-Middle Income ") + geom_smooth(method = "lm")
plot7
```

```{r}
lm1 <- lm(Disease_Burden_Index_Scaled ~ Spending_per_capita_log , data=Total22) #run regression
summary(lm1)
```

#Validation
```{r}


lm2<- cbind(Total22,lm1$residuals, lm1$fitted.values)
lm2 <- lm2 %>%
  rename(res = "lm1$residuals", fit= "lm1$fitted.values")

lm2%>%
  ggplot(aes(res)) +
  geom_histogram(bins=15, color="blue", fill="light blue")

lm2%>%
  ggplot(aes(fit, res)) +
  geom_point() 

shapiro.test(lm2$res)
qqnorm(lm2$res)


```




```{r}
lm5 <- lm(Disease_Burden_Index_Scaled ~ Spending_per_capita_log + region, data=Total22) 
summary(lm5)
```

#Filter for low income
```{r}
Total23 <- Total20 %>%
  filter(Income_Group== "Low-income")
```


#Linear model for low income group
```{r}
plot8 <- ggplot(Total23,aes( Spending_per_capita_log, Disease_Burden_Index_Scaled)) + geom_point() + labs(y= "Disease Burden Index", x= "Log GHS Low Income") + geom_smooth(method = "lm")
plot8
```

```{r}
lm1 <- lm(Disease_Burden_Index_Scaled ~ Spending_per_capita_log, data=Total23) 
summary(lm1)
```


#Validation
```{r}


lm2<- cbind(Total23,lm1$residuals, lm1$fitted.values)
lm2 <- lm2 %>%
  rename(res = "lm1$residuals", fit= "lm1$fitted.values")

lm2%>%
  ggplot(aes(res)) +
  geom_histogram(bins=15, color="blue", fill="light blue")

lm2%>%
  ggplot(aes(fit, res)) +
  geom_point() 

shapiro.test(lm2$res)
qqnorm(lm2$res)


```


```{r}
lm6 <- lm(Disease_Burden_Index_Scaled ~ Spending_per_capita_log + region, data=Total23) 
summary(lm6)
```

#Filter for upper middle income
```{r}
Total24 <- Total20 %>%
  filter(Income_Group== "Upper-middle-income")
```
#Linear model for upper middle income
```{r}
plot9 <- ggplot(Total24,aes( Spending_per_capita_log, Disease_Burden_Index_Scaled)) + geom_point() + labs(y= "Disease Burden Index", x= "Log GHS Upper-Middle Income") + geom_smooth(method = "lm")
plot9
```

```{r}
grid.arrange(plot8, plot7, plot9, plot6, nrow = 2, ncol=2)
```




```{r}
lm1 <- lm(Disease_Burden_Index_Scaled ~ Spending_per_capita_log, data=Total24) 
summary(lm1)
```



#Validation
```{r}


lm2<- cbind(Total24,lm1$residuals, lm1$fitted.values)
lm2 <- lm2 %>%
  rename(res = "lm1$residuals", fit= "lm1$fitted.values")

lm2%>%
  ggplot(aes(res)) +
  geom_histogram(bins=15, color="blue", fill="light blue")

lm2%>%
  ggplot(aes(fit, res)) +
  geom_point() 

shapiro.test(lm2$res)
qqnorm(lm2$res)


```


```{r}
lm7 <- lm(Disease_Burden_Index_Scaled ~ Spending_per_capita_log + region, data=Total24) 
summary(lm7)
```

#Filter by region

#Filter for Americas only
```{r}
Total25 <- Total20 %>%
  filter(region == "Americas")
```



```{r}
plot7 <- ggplot(Total25,aes( Spending_per_capita_log, Disease_Burden_Index_Scaled)) + geom_point() + labs(y= "Disease Burden Index", x= "Log GHS Americas") + geom_smooth(method = "lm")
plot7
```

#Linear Model for Americas region
```{r}
lm1 <- lm(Disease_Burden_Index_Scaled ~ Spending_per_capita_log, data=Total25) 
summary(lm1)
```

#Validation
```{r}


lm2<- cbind(Total25,lm1$residuals, lm1$fitted.values)
lm2 <- lm2 %>%
  rename(res = "lm1$residuals", fit= "lm1$fitted.values")

lm2%>%
  ggplot(aes(res)) +
  geom_histogram(bins=15, color="blue", fill="light blue")

lm2%>%
  ggplot(aes(fit, res)) +
  geom_point() 

shapiro.test(lm2$res)
qqnorm(lm2$res)


```






```{r}
lm8 <- lm(Disease_Burden_Index_Scaled ~ Spending_per_capita_log + Income_Group, data=Total25) 
summary(lm8)
```



#Filter for Europe
```{r}
Total26 <- Total20 %>%
  filter(region == "Europe")
```



```{r}
plot8 <- ggplot(Total26,aes( Spending_per_capita_log, Disease_Burden_Index_Scaled)) + geom_point() + labs(y= "Disease Burden Index", x= "Log GHS Europe") + geom_smooth(method = "lm")
plot8
```


#Linear model for Europe
```{r}
lm1 <- lm(Disease_Burden_Index_Scaled ~ Spending_per_capita_log, data=Total26) 
summary(lm1)
```


#Validation
```{r}


lm2<- cbind(Total26,lm1$residuals, lm1$fitted.values)
lm2 <- lm2 %>%
  rename(res = "lm1$residuals", fit= "lm1$fitted.values")

lm2%>%
  ggplot(aes(res)) +
  geom_histogram(bins=15, color="blue", fill="light blue")

lm2%>%
  ggplot(aes(fit, res)) +
  geom_point() 

shapiro.test(lm2$res)
qqnorm(lm2$res)


```




```{r}
lm8 <- lm(Disease_Burden_Index_Scaled ~ Spending_per_capita_log + Income_Group, data=Total26) 
summary(lm8)
```


#Filter for Africa
```{r}
Total27 <- Total20 %>%
  filter(region == "Africa")
```



```{r}
plot9 <- ggplot(Total27,aes( Spending_per_capita_log, Disease_Burden_Index_Scaled)) + geom_point() + labs(y= "Disease Burden Index", x= "Log GHS Africa") + geom_smooth(method = "lm")
plot9
```

```{r}
grid.arrange(plot7, plot8, plot9, ncol = 2)
```




#Africa linear model
```{r}
lm1 <- lm(Disease_Burden_Index_Scaled ~ Spending_per_capita_log, data=Total27) 
summary(lm1)
```

#Validation
```{r}


lm2<- cbind(Total27,lm1$residuals, lm1$fitted.values)
lm2 <- lm2 %>%
  rename(res = "lm1$residuals", fit= "lm1$fitted.values")

lm2%>%
  ggplot(aes(res)) +
  geom_histogram(bins=15, color="blue", fill="light blue")

lm2%>%
  ggplot(aes(fit, res)) +
  geom_point() 

shapiro.test(lm2$res)
qqnorm(lm2$res)


```



```{r}
lm8 <- lm(Disease_Burden_Index_Scaled ~ Spending_per_capita_log + Income_Group, data=Total27) 
summary(lm8)
```







#Searching for indictors of healthcare systems (private vs public sector or universal)


```{r}
national<- codes[grepl("UHC", codes$display), ] 
```

```{r}
national_df <- get_data("HSS_UHCLEGISLATION")  #create UHC dataset


```



```{r}
national_df = rename(national_df, Universal_Legislation = value) #rename columns
```

```{r}
Total31<- merge(Total20 , national_df, by= "country") #merge mortality index into total df
```


```{r}
Total31 <- subset(Total31, select = c(country, Income_Group, year.x, region.x, Spending_per_capita, GHE_percent, Adult_mort,HIV_AIDS, IHD, Neonatal, Stroke, TB_resp, Diarrhoea, Road_Injuries, birth_defects, Malaria, pneumonia, Lower_Respiratory, Neonatal_toscale, Birth_Defects_toscale, population_1000, Disease_Burden_Index_Scaled, Spending_per_capita_log, Universal_Legislation))    #select desired columns with UHC included 
```



```{r}  
Total31 = rename(Total31, year = year.x)   #rename
Total31 = rename(Total31, region = region.x) 


```

#Export updated final dataset with UHC
```{r}
write.csv(Total31, "Final_Data.csv")
```


#Universal Health Care yes/no count
```{r}
hist7 <-ggplot(data=Total20, aes(Total31$Universal_Legislation)) + 
  geom_bar(stat="count", fill = "blue") + labs(x= "Passed Universal Healthcare Legislation", y = "Count")
hist7
```



#Group by UHC and count
```{r}
TotalC <- Total31 %>%
  group_by(Universal_Legislation)%>%
  tally()
```



#Boxplot compared UHC by spending per capita
```{r}
p10 <- ggplot(Total31, aes(x = Universal_Legislation, y = Spending_per_capita)) +
        geom_boxplot(fill= "light blue") + labs (x= "Universal Legislation" , y= "Healthcare Spending per Capita (USD)")
        

p10

```

```{r}
p11 <- ggplot(Total31, aes(x = Universal_Legislation, y = Adult_mort)) +
        geom_boxplot(fill= "light blue") + labs (x= "Universal Legislation", y= "Adult Mortality per 1000 people")
        

p11

```

```{r}
p12 <- ggplot(Total31, aes(x = Universal_Legislation, y = Disease_Burden_Index_Scaled)) +
        geom_boxplot(fill= "light blue") + labs (x= "Universal Legislation",  y= "Disease Burden Index")
        

p12
```

```{r}
grid.arrange(p10, p11, p12, nrow = 1) 
```

```{r}
library(dplyr)
```


```{r}
UHC_YES <- Total31 %>%
  filter(Universal_Legislation == "Yes")

median(UHC_YES$Spending_per_capita)
median(UHC_YES$Adult_mort)
median(UHC_YES$Disease_Burden_Index_Scaled)
```


```{r}
UHC_NO <- Total31 %>%
  filter(Universal_Legislation == "No")

median(UHC_NO$Spending_per_capita)
median(UHC_NO$Adult_mort)
median(UHC_NO$Disease_Burden_Index_Scaled)
```



#T.test comparing UHC vs spending per capita
```{r}
t.test(Spending_per_capita ~ Universal_Legislation, data = Total31)
```

#Filter by UHC = Yes
```{r}
TotalD <- Total31 %>%
  filter(Universal_Legislation == "Yes")

range(TotalD$Adult_mort)
```


#Filter by UHC = no
```{r}
TotalD <- Total31 %>%
  filter(Universal_Legislation == "No")

range(TotalD$Adult_mort)
```

#Find outliers
```{r}
TotalD <- Total31 %>%
  filter(Adult_mort >400, Universal_Legislation == "No")
```


#Africa Dataset
```{r}
Total_Africa <- Total31 %>%
  filter(region== "Africa")

mean(Total_Africa$Disease_Burden_Index_Scaled)

Total_Africa_count <- Total_Africa %>%
  group_by(Universal_Legislation)%>%
  tally()
  
```


#Filter by UHC = NO
```{r}
Total_Africa_mean <- Total_Africa %>%
  filter(Universal_Legislation== "No")
mean(Total_Africa_mean$Disease_Burden_Index_Scaled)
```

#Find outliers
```{r}
Total_Africa_out <- Total_Africa %>%
  filter(Universal_Legislation == "No", Disease_Burden_Index_Scaled < .3)
```


#T.test with Disease burden vs UHC in Africa alone
```{r}
t.test(Disease_Burden_Index_Scaled ~ Universal_Legislation, data = Total_Africa)
```

#Boxplot with UHC vs Disease burden in Africa alone
```{r}
p10 <- ggplot(Total_Africa, aes(x = Universal_Legislation, y = Disease_Burden_Index_Scaled)) +
        geom_boxplot(fill= "red") + labs (x= "Universal Legislation in Africa",  y= "Disease Burden Index in African Countries")
        

p10

```


#Europe Dataset
```{r}
Total_Europe <- Total31 %>%
  filter(region== "Europe")

mean(Total_Europe$Disease_Burden_Index_Scaled)
  
Total_Europe_count <- Total_Europe %>%
  group_by(Universal_Legislation)%>%
  tally()
```

```{r}
Total_Europe_mean <- Total_Europe %>%
  filter(Universal_Legislation== "Yes")
mean(Total_Europe_mean$Disease_Burden_Index_Scaled)
```

```{r}
Total_Europe_YES <- Total_Europe %>%
  filter(Universal_Legislation == "Yes")
Total_Europe_No <- Total_Europe %>%
  filter(Universal_Legislation == "No")
Total_Africa_No <- Total_Africa %>%
  filter(Universal_Legislation == "No")
Total_Africa_YES <- Total_Africa %>%
  filter(Universal_Legislation == "Yes")

median(Total_Europe_YES$Disease_Burden_Index_Scaled)
median(Total_Africa_YES$Disease_Burden_Index_Scaled)
median(Total_Europe_No$Disease_Burden_Index_Scaled)
median(Total_Africa_No$Disease_Burden_Index_Scaled)

```



#t.test with UHC vs disease burden in Europe alone
```{r}
t.test(Disease_Burden_Index_Scaled ~ Universal_Legislation, data = Total_Europe)
```

```{r}
p11 <- ggplot(Total_Europe, aes(x = Universal_Legislation, y = Disease_Burden_Index_Scaled)) +
        geom_boxplot(fill= "blue") + labs (x= "Universal Legislation in Europe",  y= "Disease Burden Index in European Countries")
        

p11

```

```{r}
grid.arrange(p10, p11, nrow = 1) 
```



```{r}
Total_Europe_out <- Total_Europe %>%
  filter(Universal_Legislation == "Yes", Disease_Burden_Index_Scaled > .3)
```





#Disease burden vs UHC box plot, total dataset
```{r}
p10 <- ggplot(Total31, aes(x = Universal_Legislation, y = Disease_Burden_Index_Scaled)) +
        geom_boxplot(fill= "light blue") + labs (x= "Passed Universal Healthcare Legislation",  y= "Disease Burden Index")
        

p10

```

#T test with DBI vs UHC, total dataset
```{r}
t.test(Disease_Burden_Index_Scaled ~ Universal_Legislation, data = Total31)
```



#Filter by income group
```{r}
Total_LowIncome <- Total31 %>%
  filter(Income_Group== "Low-income" | Income_Group == "Lower-middle-income")

mean(Total_LowIncome$Disease_Burden_Index_Scaled)
```

#Boxplot and t test with low income countries only: UHC vs DBI
```{r}
p10 <- ggplot(Total_LowIncome, aes(x = Universal_Legislation, y = Disease_Burden_Index_Scaled)) +
        geom_boxplot(fill= "purple") + labs (x= " Low & Lower Middle Income",  y= "Disease Burden Index")
        

p10

```

```{r}
t.test(Disease_Burden_Index_Scaled ~ Universal_Legislation, data = Total_LowIncome)
```

```{r}
Total_LowIncome_count <- Total_LowIncome %>%
  group_by(Universal_Legislation) %>%
  tally()
```



#High income only
```{r}
Total_HighIncome <- Total31 %>%
  filter(Income_Group == "High-income" )

mean(Total_HighIncome$Disease_Burden_Index_Scaled)
```

```{r}
p11 <- ggplot(Total_HighIncome, aes(x = Universal_Legislation, y = Disease_Burden_Index_Scaled)) +
        geom_boxplot(fill= "light green") + labs (x= "High Income",  y= "Disease Burden Index")
        

p11

```

```{r}
t.test(Disease_Burden_Index_Scaled ~ Universal_Legislation, data = Total_HighIncome)
```

```{r}
Total_HighIncome_count <- Total_HighIncome %>%
  group_by(Universal_Legislation) %>%
  tally()
```

#Middle Income Only
```{r}
Total_MiddleIncome <- Total31 %>%
  filter(Income_Group == "Upper-middle-income" )


```


```{r}
p12 <- ggplot(Total_MiddleIncome, aes(x = Universal_Legislation, y = Disease_Burden_Index_Scaled)) +
        geom_boxplot(fill= "light pink") + labs (x= "Upper Middle Income",  y= "Disease Burden Index")
        

p12

```

```{r}
grid.arrange(p10, p12, p11, nrow = 1) 
```



```{r}
t.test(Disease_Burden_Index_Scaled ~ Universal_Legislation, data = Total_MiddleIncome)
```

```{r}
Total_MiddleIncome_count <- Total_MiddleIncome %>%
  group_by(Universal_Legislation) %>%
  tally()
```


```{r}
Total_LowIncome_YES <- Total_LowIncome %>%
  filter(Universal_Legislation== "Yes")
Total_LowIncome_No <- Total_LowIncome %>%
  filter(Universal_Legislation== "No")



median(Total_LowIncome_YES$Disease_Burden_Index_Scaled)
median(Total_LowIncome_No$Disease_Burden_Index_Scaled)
```


```{r}
Total_Middle_YES <- Total_MiddleIncome %>%
  filter(Universal_Legislation== "Yes")
Total_MiddleIncome_No <- Total_MiddleIncome %>%
  filter(Universal_Legislation== "No")



median(Total_Middle_YES$Disease_Burden_Index_Scaled)
median(Total_MiddleIncome_No $Disease_Burden_Index_Scaled)
```


```{r}
Total_High_YES <- Total_HighIncome %>%
  filter(Universal_Legislation== "Yes")
Total_HighIncome_No <- Total_HighIncome %>%
  filter(Universal_Legislation== "No")



median(Total_High_YES$Disease_Burden_Index_Scaled)
median(Total_HighIncome_No$Disease_Burden_Index_Scaled)
```


